---
title: "Aggregation"
description: >
  How to use the included aggregators, build custom ones, and aggregate the instrument status log.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Aggregation}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

Raw files differ in which information they provide depending on the instrument and its software version. `orbi_aggregate_raw()` therefore uses an **aggregator** that defines which information to pull out of the raw files, where to find it, and what data type it should have. This vignette shows how to use the included aggregators, how to build your own, and how to use a custom aggregator to retrieve information from the instrument status log.

```{r, include=FALSE}
# default chunk options
knitr::opts_chunk$set(collapse = FALSE, message = TRUE, comment = "")
```

```{r, message=FALSE}
# libraries
library(isoorbi) # load isoorbi R package
library(dplyr) # for mutating data frames
library(ggplot2) # for data visualization
```

# Example files

Here we use two example files from two different instruments (an Orbitrap Exploris 240 and a Q Exactive Focus).

```{r}
# two example files from different instruments
raw_files <-
  orbi_get_example_files(c("dual_inlet.raw", "s3744.RAW")) |>
  orbi_read_raw()
raw_files
```

# Included aggregators

## `orbi_get_aggregator()`

`orbi_aggregate_raw()` uses the `standard` aggregator by default. The `minimal` aggregator contains a smaller set of columns to aggregate. The `extended` aggregator is more elaborate, providing access to additional columns from the raw data files.

```{r}
# example: minimal vs. extended aggregator
orbi_get_aggregator("minimal")
orbi_get_aggregator("extended")
```

```{r}
# using the extended aggregator instead of the default (standard)
raw_files |> orbi_aggregate_raw(aggregator = "extended")
```

The printout of the aggregated data only says how many columns of each dataset were *not aggregated*. To see which ones they are (e.g. to find additional information to include in a custom aggregator), print the aggregated data with `show_all = TRUE`.

```{r}
agg_data <- raw_files |> orbi_aggregate_raw()
agg_data |> print(show_all = TRUE)
```

# Custom aggregators

## `orbi_register_aggregator()`

You can build your own aggregator with `orbi_start_aggregator()` and/or expand an existing one with `orbi_add_to_aggregator()` and then register it via `orbi_register_aggregator()` to use it by name.

```{r}
my_agg <-
  orbi_get_aggregator("minimal") |>
  # pull out the S-Lens RF Level information from the scans and store it as a number
  orbi_add_to_aggregator(
    "scans",
    "slens_rf",
    source = "S-Lens RF Level",
    cast = "as.numeric"
  ) |>
  orbi_register_aggregator(name = "custom")

# show my aggregator
my_agg

# use it
raw_files |> orbi_aggregate_raw(aggregator = "custom")
```

# Instrument status log

Besides the scans, raw files hold the instrument's **status log**: readbacks that the instrument records independently of the scans (typically every few seconds) such as temperatures, pressures, voltages, and ion source settings. `orbi_read_raw()` returns it as the `status_log` dataset.

```{r}
# the raw status log of the first file (one row per log entry, one column per channel)
raw_files$status_log[[1]]
```

Which channels a status log contains depends entirely on the instrument, so none of the included aggregators take anything from it. Instead, `orbi_aggregate_raw()` reports all of the status log channels as *not aggregated* (see `print(agg_data, show_all = TRUE)` above) so you can pick the ones you are interested in.

## Status log aggregator

The status log stores all channel values as text (exactly as the instrument reports them), so the aggregator needs to `cast` the channels you want to use to numbers. Because different instruments can call the same channel by different names, you can provide the alternative names as the `source` and the aggregator uses whichever one exists in each file. Channels that a file does not have at all are left empty (`NA`) for that file.

```{r}
status_agg <-
  orbi_get_aggregator("standard") |>
  # when each status log entry was recorded
  orbi_add_to_aggregator(
    "status_log",
    "time.min",
    source = "StartTime",
    cast = "as.numeric"
  ) |>
  # ion source
  orbi_add_to_aggregator(
    "status_log",
    "spray_voltage.V",
    source = "Spray Voltage (V)",
    cast = "as.numeric"
  ) |>
  orbi_add_to_aggregator(
    "status_log",
    "spray_current.uA",
    source = "Spray Current (µA)",
    cast = "as.numeric"
  ) |>
  # temperatures (some channels have different names on the two instruments)
  orbi_add_to_aggregator(
    "status_log",
    "ambient_temp.C",
    source = c("Ambient temp. (°C)", "Ambient Temperature (°C)"),
    cast = "as.numeric"
  ) |>
  orbi_add_to_aggregator(
    "status_log",
    "orbitrap_temp.C",
    source = c("Orbitrap block temp. (°C)", "Analyzer Temperature (°C)"),
    cast = "as.numeric"
  ) |>
  orbi_add_to_aggregator(
    "status_log",
    "detector_temp.C",
    source = "Detector temp. (°C)",
    cast = "as.numeric"
  ) |>
  # gas pressures
  orbi_add_to_aggregator(
    "status_log",
    "sheath_gas.psi",
    source = "SB: Sheath gas pres. (psi)",
    cast = "as.numeric"
  ) |>
  orbi_add_to_aggregator(
    "status_log",
    "aux_gas.psi",
    source = "SB: Aux. gas pres. (psi)",
    cast = "as.numeric"
  ) |>
  orbi_register_aggregator("standard with status log")

# show summary
status_agg

# aggregate with the status log aggregator
status_agg_data <- raw_files |>
  orbi_aggregate_raw(aggregator = "standard with status log")
```

## Visualizing the status log

Status log data is a great way to check on the instrument conditions during an analysis, for example to see whether the ion source and the temperatures were stable.

```{r}
#| label: fig-status-log
#| fig-cap: Ion source settings, temperatures, and gas pressures recorded in the status logs of the two files.
#| fig-width: 8
#| fig-height: 12

# retrieve the status log channels
status_log <- status_agg_data |>
  orbi_get_data(file_info = "filename", status_log = everything())
status_log

# bring the channels into long format for plotting them in panels
status_log |>
  tidyr::pivot_longer(
    cols = -c("uidx", "filename", "time.min"),
    names_to = "channel"
  ) |>
  # channels in the order they were defined in the aggregator
  mutate(channel = factor(channel, levels = unique(channel))) |>
  # filter out channels a file does not have
  filter(!is.na(value)) |>
  ggplot() +
  aes(x = time.min, y = value) +
  geom_line() +
  facet_grid(channel ~ filename, scales = "free") +
  labs(x = "time [min]", y = NULL) +
  theme_bw()
```

## Comparison with the scans

The scans also record the temperature of the Orbitrap analyzer (`analyzerTemperature`, aggregated by the `standard` aggregator), so we can compare it with the temperatures in the status log: the ambient temperature, the Orbitrap block (analyzer) temperature, and the detector temperature. Note that the analyzer temperature in the scans is not identical to the Orbitrap temperature in the status log: it is not as precise and differs from the status log readback by up to ~0.2 °C in these files. In the longer run below (`dual_inlet`), it behaves like a strongly smoothed version of the status log's Orbitrap block temperature that lags behind its changes. We are not sure yet why these are different.

```{r}
#| label: fig-temperatures
#| fig-cap: Temperatures recorded in the status log and the analyzer (Orbitrap) temperature recorded in the scans.
#| fig-width: 8
#| fig-height: 4

# temperatures from the status log
temps <- status_agg_data |>
  orbi_get_data(
    file_info = "filename",
    status_log = c(
      "time.min",
      "ambient" = "ambient_temp.C",
      "Orbitrap" = "orbitrap_temp.C",
      "detector" = "detector_temp.C"
    )
  ) |>
  tidyr::pivot_longer(
    cols = c("ambient", "Orbitrap", "detector"),
    names_to = "temperature",
    values_to = "temperature.C"
  ) |>
  mutate(data_source = "status log") |>
  # the analyzer (Orbitrap) temperature from the scans
  bind_rows(
    status_agg_data |>
      orbi_get_data(
        file_info = "filename",
        scans = c("time.min", "temperature.C" = "analyzerTemperature")
      ) |>
      mutate(temperature = "Orbitrap", data_source = "scans")
  ) |>
  # temperatures a file does not have
  filter(!is.na(temperature.C))

# compare
temps |>
  ggplot() +
  aes(
    x = time.min,
    y = temperature.C,
    color = temperature,
    linetype = data_source
  ) +
  geom_line() +
  facet_wrap(~filename, scales = "free") +
  scale_linetype_manual(values = c(3, 1)) +
  scale_color_brewer(palette = "Dark2") +
  labs(x = "time [min]", y = "temperature [°C]") +
  theme_bw()
```

