---
title: "Functionality Guide"
description: >
  A detailed guide to isoorbi functionality.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Functionality Guide}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

> This step-by-step functionality guide is still in development. Eventually all functions in the [package structure flowchart](https://isoorbi.isoverse.org/index.html#package-structure) will be covered with detailed examples. All functions below labelled with an `*` are required steps of the standard data processing flow. Everything else is optional. Rarely used additional features that are mentioned here but not part of the standard flowchart are labeled as `bonus`.

```{r, include=FALSE}
# default chunk options
knitr::opts_chunk$set(collapse = FALSE, message = TRUE, comment = "")
```


```{r, message=FALSE}
# libraries
library(isoorbi) #load isoorbi R package
library(dplyr) # for mutating data frames
library(ggplot2) # for data visualization
```


# Reading raw files

First step is reading in your .raw data files.

## `orbi_find_raw()`

```{r}
#| include: false
if (!dir.exists("data")) {
  dir.create("data")
}
system.file(package = "isoorbi", "extdata") |>
  orbi_find_raw() |>
  paste0(".cache.zip") |>
  file.copy(to = "data", overwrite = TRUE)
```

```{r}
#| label: read files
# path to your data folder
data_folder <- file.path("data")

# finding raw files with "nitrate" in the name in the data folder
file_paths <- data_folder |> orbi_find_raw(pattern = "nitrate")

# show what was found
file_paths
```

## `orbi_read_raw()` *

```{r}
# read files (simplest)
raw_files <- file_paths |> orbi_read_raw()

# read files including some raw spectra
raw_files <-
  file_paths |>
  # load the spectra from scans 1, 10, and 100
  orbi_read_raw(include_spectra = c(1, 10, 100)) |>
  # you can quiet any call with suppressMessages
  # that way it won't print the info message
  suppressMessages()

# show summary for the files that were read
raw_files
```

## `orbi_aggregate_data()` *

Combine (aggregate) the data from the raw files.

```{r}
# aggregate raw data
agg_data <- raw_files |> orbi_aggregate_raw()

# show all that was recovered
agg_data

# also show all that was ignored/not aggregated
agg_data |> print(show_all = TRUE)
```

For how to use the other included aggregators, how to build your own, and how to aggregate the instrument status log, see the [aggregation vignette](aggregation.html).

## bonus `orbi_get_problems()`

There were no problems reading and/or aggregating the raw data so these are empty but this can be very helpful to see what went wrong during reading or aggregation.

```{r}
raw_files |> orbi_get_problems()
agg_data |> orbi_get_problems()
```

## `orbi_get_data()`

At this point (and any later point), you can always extract the data of interest from the aggregated data set using `orbi_get_data()`. If you prefer working with a data frame tibble from `orbi_get_data()` instead of the aggregated data structure, you can switch to that at any point and use the resulting data frame tibble in subsequent functions.

```{r}
# direct access to the data stored in the aggregated dataset
agg_data$file_info
agg_data$scans
agg_data$peaks
agg_data$spectra
agg_data$status_log

# better way to retrieve+combine the data with dplyr select syntax:
agg_data |>
  orbi_get_data(
    file_info = c(
      "filename",
      "creation_date",
      "instrument" = "InstrumentModel"
    ),
    scans = c("time.min", "tic", "resolution"),
    peaks = c("mz" = "mzMeasured", starts_with("peak"))
  )
```

# Identifying isotopocules

The next step is identifying isotpocules.

## `orbi_identify_isotopocules()` *

```{r}
# list of isotopocules (can alternatively be in a tsv/csv/xlsx file)
isotopocules <- tibble(
  compound = "nitrate",
  isotopolog = c("M0", "15N", "17O", "18O"),
  mass = c(61.9878, 62.9850, 62.9922, 63.9922),
  tolerance = 1,
  charge = 1
)

# identify
data <- agg_data |> orbi_identify_isotopocules(isotopocules)
```

## `orbi_plot_spectra()`

# Data checks

## `orbi_flag_satellite_peaks()` *

```{r}
#| label: fig-satellite-peaks
#| fig-cap: Satellite peaks in the 10 scan test file.
#| fig-width: 6
#| fig-height: 4
#| warning: false
# this can happen here or later on in the workflow
# in the case of these files there are no satellite peaks
data |>
  orbi_filter_files("nitrate_test_10scans") |>
  orbi_flag_satellite_peaks() |>
  orbi_plot_satellite_peaks()
```

## `orbi_plot_isotopocule_coverage()`

```{r}
#| label: fig-coverage
#| fig-cap: Isotopocule coverage of the 10 scan test file.
#| fig-width: 8
#| fig-height: 6

# this can happen here or later on in the workflow
data |> orbi_get_isotopocule_coverage()
data |>
  orbi_filter_files("nitrate_test_10scans") |>
  orbi_plot_isotopocule_coverage()
```

# Data blocks

## `orbi_define_blocks()`

Data blocks mark which scans belong together (e.g. one sample in a flow injection) and which scans are not used for the ratio calculations. Each block is defined either by time (`start_time.min` and `end_time.min`) or by scan number (`start_scan.no` and `end_scan.no`). By default, a block is added to every file. Note that to let a block stretch until the last data point, use `Inf` for `end_time.min` or `end_scan.no`. Use `orbi_get_blocks_info()` to get a table of the defined blocks.


```{r}
# one block for all files (by time)
data |>
  orbi_define_blocks(start_time.min = 0, end_time.min = 0.015) |>
  orbi_get_blocks_info()
```

To add a block only to specific file(s), use `in_filename`. Like all the other parameters, it can be a single value or a vector with one value per block. Use `orbi_plot_raw_data()` to visualize where blocks have been defined.

```{r}
#| label: fig-blocks
#| fig-cap: The two data blocks defined in the 10 scan test file only.
#| fig-width: 8
#| fig-height: 4
#| message: false
data |>
  orbi_define_blocks(
    start_scan.no = c(1, 6),
    end_scan.no = c(4, Inf),
    block_name = c("start", "finish"),
    in_filename = "nitrate_test_10scans"
  ) |>
  # plot with scans instead of time as x axis and show all data points for clarity
  orbi_plot_raw_data(y = tic, x = "scan.no", show_points = TRUE)
```

For defining many blocks in a file, it is often easier to provide them as a `blocks_table` (which could also come from a spreadsheet) with one row per block. An `in_filename` of `NA` adds that block to all files.

```{r}
#| label: fig-blocks2
#| fig-cap: Data blocks defined from a blocks table and shown with the block names.
#| fig-width: 8
#| fig-height: 4
#| message: false
blocks_table <- tibble(
  start_scan.no = c(1, 3, 6),
  end_scan.no = c(1, 4, Inf),
  block_name = c("first", "second", "end"),
  in_filename = c(NA, "nitrate_test_10scans", "nitrate_test_10scans")
)
blocks_table

data |>
  orbi_define_blocks(blocks_table = blocks_table) |>
  orbi_plot_raw_data(
    y = tic * it.ms,
    x = "scan.no",
    show_points = TRUE,
    # to show the names of the data blocks, turn this setting on
    use_data_block_names = TRUE
  )
```

# Ratio calculations

## `orbi_define_basepeak()` *

## `orbi_summarize_results()` *