---
title: "Functionality Guide"
description: >
  A detailed guide to isoorbi functionality.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Functionality Guide}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

> This step-by-step functionality guide is still in development. Eventually all functions in the [package structure flowchart](https://isoorbi.isoverse.org/index.html#package-structure) will be covered with detailed examples. All functions below labelled with an `*` are required steps of the standard data processing flow. Everything else is optional. Rarely used additional features that are mentioned here but not part of the standard flowchart are labeled as `bonus`.

```{r, include=FALSE}
# default chunk options
knitr::opts_chunk$set(collapse = FALSE, message = TRUE, comment = "")
```


```{r, message=FALSE}
# libraries
library(isoorbi) #load isoorbi R package
library(dplyr) # for mutating data frames
library(ggplot2) # for data visualization
```


# Reading raw files

First step is reading in your .raw data files.

## `orbi_find_raw()`

```{r}
#| include: false
if (!dir.exists("data")) {
  dir.create("data")
}
system.file(package = "isoorbi", "extdata") |>
  orbi_find_raw() |>
  paste0(".cache.zip") |>
  file.copy(to = "data", overwrite = TRUE)
```

```{r}
#| label: read files
# path to your data folder
data_folder <- file.path("data")

# finding raw files with "nitrate" in the name in the data folder
file_paths <- data_folder |> orbi_find_raw(pattern = "nitrate")

# show what was found
file_paths
```

## `orbi_read_raw()` *

```{r}
# read files (simplest)
raw_files <- file_paths |> orbi_read_raw()

# read files including some raw spectra
raw_files <-
  file_paths |>
  # load the spectra from scans 1, 10, and 100
  orbi_read_raw(include_spectra = c(1, 10, 100)) |>
  # you can quiet any call with suppressMessages
  # that way it won't print the info message
  suppressMessages()

# show summary for the files that were read
raw_files
```

## `orbi_aggregate_data()` *

Combine (aggregate) the data from the raw files.

```{r}
# aggregate raw data
agg_data <- raw_files |> orbi_aggregate_raw()

# show all that was recovered
agg_data

# also show all that was ignored/not aggregated
agg_data |> print(show_all = TRUE)
```

### bonus `orbi_get_aggregator()`

You can optionally use a different aggregator. The `minimal` aggregator contains a smaller set of columns to aggregate. The `extended` aggregator is more elaborate, providing access to additional columns from the raw data files.

```{r}
# example: minimal vs. extended aggregator
orbi_get_aggregator("minimal")
orbi_get_aggregator("extended")

# using the extended aggregator instead of the default (standard)
raw_files |> orbi_aggregate_raw(aggregator = "extended")
```

### bonus `orbi_register_aggregator()`

Or even build your own aggregator with `orbi_start_aggregator()` and/or expand an existing one with `orgi_add_to_aggregator()` and then register it via `orbi_register_aggregator()`. This funnctionality is rarely needed and thus not part of the package structure flowchart.

```{r}
my_agg <-
  orbi_get_aggregator("minimal") |>
  # pull out the S-Lens RF Level information from the scans and store it as a number
  orbi_add_to_aggregator(
    "scans",
    "slens_rf",
    source = "S-Lens RF Level",
    cast = "as.numeric"
  ) |>
  orbi_register_aggregator(name = "test")

# show my agg summary
my_agg

# use it
raw_files |> orbi_aggregate_raw(aggregator = "test")
```

## bonus `orbi_get_problems()`

There were no problems reading and/or aggregating the raw data so these are empty but this can be very helpful to see what went wrong during reading or aggregation.

```{r}
raw_files |> orbi_get_problems()
agg_data |> orbi_get_problems()
```

## `orbi_get_data()`

At this point (and any later point), you can always extract the data of interest from the aggregated data set using `orbi_get_data()`. If you prefer working with a data frame tibble from `orbi_get_data()` instead of the aggregated data structure, you can switch to that at any point and use the resulting data frame tibble in subsequent functions.

```{r}
# direct access to the data stored in the aggregated dataset
agg_data$file_info
agg_data$scans
agg_data$peaks
agg_data$spectra
agg_data$status_log

# better way to retrieve+combine the data with dplyr select syntax:
agg_data |>
  orbi_get_data(
    file_info = c(
      "filename",
      "creation_date",
      "instrument" = "InstrumentModel"
    ),
    scans = c("time.min", "tic", "resolution"),
    peaks = c("mz" = "mzMeasured", starts_with("peak"))
  )
```

# Bonus: working with the `status_log`

Besides the scans, raw files hold the instrument's **status log**: readbacks that the instrument records independently of the scans (typically every few seconds) such as temperatures, vacuum pressures, voltages, and ion source settings. `orbi_read_raw()` returns it as the `status_log` dataset. Which channels a status log contains depends entirely on the instrument, so none of the included aggregators take anything from it. Instead, `orbi_aggregate_raw()` reports the available status log channels as *not aggregated* so you can pick the ones you are interested in. Printing the aggregated data only says how many there are, use `print(agg_data, show_all = TRUE)` to list all of them.

Here we use two example files from two different instruments (an Orbitrap Exploris 240 and a Q Exactive Focus) to show how to aggregate a few channels of interest.

```{r}
# two example files from different instruments
status_files <-
  orbi_get_example_files(c("dual_inlet.raw", "s3744.RAW")) |>
  orbi_read_raw()

# the raw status log of the first file (one row per log entry, one column per channel)
status_files$status_log[[1]]
```

The status log stores all channel values as text (exactly as the instrument reports them), so the aggregator needs to `cast` the channels you want to use to numbers. Because the two instruments call the same channels by different names, you can provide the alternative names as the `source` and the aggregator uses whichever one exists in each file.

```{r}
status_agg <-
  orbi_get_aggregator("standard") |>
  # when each status log entry was recorded
  orbi_add_to_aggregator(
    "status_log",
    "time.min",
    source = "StartTime",
    cast = "as.numeric"
  ) |>
  # temperatures (the channel names differ between the instruments)
  orbi_add_to_aggregator(
    "status_log",
    "ambient_temp.C",
    source = c("Ambient temp. (°C)", "Ambient Temperature (°C)"),
    cast = "as.numeric"
  ) |>
  orbi_add_to_aggregator(
    "status_log",
    "analyzer_temp.C",
    source = c("Orbitrap block temp. (°C)", "Analyzer Temperature (°C)"),
    cast = "as.numeric"
  ) |>
  # ultra high vacuum pressure
  orbi_add_to_aggregator(
    "status_log",
    "uhv_pressure.mbar",
    source = c("ICB: UHV pres. (mbar)", "UHV Sensor (mbar)"),
    cast = "as.numeric"
  ) |>
  orbi_register_aggregator("standard with status logs")

# show summary
status_agg

# aggregate with the status log aggregator
status_agg_data <- status_files |>
  orbi_aggregate_raw(aggregator = "standard with status logs")
```

Status log data is a great way to check on the instrument conditions during an analysis, for example to see whether the temperatures were stable or whether there were any vacuum issues.

```{r}
#| label: fig-status-log
#| fig-cap: Instrument temperatures and vacuum pressure recorded in the status logs of the two files.
#| fig-width: 8
#| fig-height: 7

# retrieve the status log channels
status_log <- status_agg_data |>
  orbi_get_data(file_info = "filename", status_log = everything())

# bring the channels into long format for plotting them in panels
status_log |>
  tidyr::pivot_longer(
    cols = c("ambient_temp.C", "analyzer_temp.C", "uhv_pressure.mbar"),
    names_to = "channel"
  ) |>
  ggplot() +
  aes(x = time.min, y = value) +
  geom_line() +
  facet_grid(channel ~ filename, scales = "free") +
  labs(x = "time [min]", y = NULL) +
  theme_bw()
```

# Identifying isotopocules

The next step is identifying isotpocules.

## `orbi_identify_isotopocules()` *

```{r}
# list of isotopocules (can alternatively be in a tsv/csv/xlsx file)
isotopocules <- tibble(
  compound = "nitrate",
  isotopolog = c("M0", "15N", "17O", "18O"),
  mass = c(61.9878, 62.9850, 62.9922, 63.9922),
  tolerance = 1,
  charge = 1
)

# identify
data <- agg_data |> orbi_identify_isotopocules(isotopocules)
```

## `orbi_plot_spectra()`

# Data checks

## `orbi_flag_satellite_peaks()` *

```{r}
#| label: fig-satellite-peaks
#| fig-cap: Satellite peaks in the 10 scan test file.
#| fig-width: 6
#| fig-height: 4
#| warning: false
# this can happen here or later on in the workflow
# in the case of these files there are no satellite peaks
data |>
  orbi_filter_files("nitrate_test_10scans") |>
  orbi_flag_satellite_peaks() |>
  orbi_plot_satellite_peaks()
```

## `orbi_plot_isotopocule_coverage()`

```{r}
#| label: fig-coverage
#| fig-cap: Isotopocule coverage of the 10 scan test file.
#| fig-width: 8
#| fig-height: 6

# this can happen here or later on in the workflow
data |> orbi_get_isotopocule_coverage()
data |>
  orbi_filter_files("nitrate_test_10scans") |>
  orbi_plot_isotopocule_coverage()
```

# Data blocks

## `orbi_define_blocks()`

Data blocks mark which scans belong together (e.g. one sample in a flow injection) and which scans are not used for the ratio calculations. Each block is defined either by time (`start_time.min` and `end_time.min`) or by scan number (`start_scan.no` and `end_scan.no`). By default, a block is added to every file.

```{r}
# one block for all files (by time)
data |>
  orbi_define_blocks(start_time.min = 0, end_time.min = 0.015) |>
  orbi_get_blocks_info()
```

Files often differ in length. To let a block last until the end of each file, use `Inf` for `end_time.min` or `end_scan.no`.

```{r}
# one block from the first until the last scan of each file
data |>
  orbi_define_blocks(start_scan.no = 1, end_scan.no = Inf) |>
  orbi_get_blocks_info()
```

To add a block only to specific file(s), use `in_filename`. Like all the other parameters, it can be a single value or a vector with one value per block.

```{r}
# two blocks (by scan number) only in the 10 scan file
blocks_data <- data |>
  orbi_define_blocks(
    start_scan.no = c(2, 6),
    end_scan.no = c(4, 9),
    block_name = c("first", "second"),
    in_filename = "nitrate_test_10scans"
  )

# overview of the blocks in each file
blocks_data |> orbi_get_blocks_info()
```

```{r}
#| label: fig-blocks
#| fig-cap: The two blocks defined in the 10 scan test file (scans outside the blocks are unused).
#| fig-width: 8
#| fig-height: 4
blocks_data |>
  orbi_filter_files("nitrate_test_10scans") |>
  orbi_plot_raw_data(y = tic * it.ms, x = "scan.no") +
  geom_point(size = 4) # add the data points for clarity
```

For many blocks, it is often easier to provide them as a `blocks_table` (which could also come from a spreadsheet) with one row per block. An `in_filename` of `NA` adds that block to all files.

```{r}
blocks_table <- tibble(
  start_scan.no = c(1, 2, 6),
  end_scan.no = c(1, 4, 9),
  block_name = c("all", "first", "second"),
  in_filename = c(NA, "nitrate_test_10scans", "nitrate_test_10scans")
)
blocks_table

data |>
  orbi_define_blocks(blocks_table = blocks_table) |>
  orbi_get_blocks_info()
```

# Ratio calculations

## `orbi_define_basepeak()` *

## `orbi_summarize_results()` *